Sequential algorithms to split and merge ultra-high resolution 3D images

### 2017-08-03 paper.tex
Worked on results

% packages
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amsfonts} % for R symbol (the set of real numbers)
\usepackage{color}
\usepackage[pdftex]{graphicx}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=black,citecolor=black,filecolor=black,urlcolor=blue}
\usepackage{mathtools}
\usepackage{multirow}
\usepackage{stmaryrd} % for llbracket and rrbracket
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
\DeclarePairedDelimiter{\floor}{\lfloor}{\rfloor}

@@

This paper makes the following contributions:
\begin{itemize}
  \item We propose a set of algorithms to split and merge
    terabyte-scale 3D images from blocks or slices.
  \item We model the performance (number of seeks) of these algorithms.
  \item We evaluate our model and algorithms on a representative
    use-case involving a 1~TB image a hard drive and a solid-state
    drive.
\end{itemize}
Section~\ref{sec:algos} presents our algorithms and their complexity,
Section~\ref{sec:results} details the results and
Sections~\ref{sec:discussion} discusses them.\todo{Revise when outline is final.}

@@

\subsection{Lossless compression}

Lossless compression might reduce overall read and write times by
reducing seek time and data sizes. However, it needs to be implemented
on-the-fly, i.e., in chunks. See the work in~\cite{rajna2015speeding}
and
\href{https://github.com/pauldmccarthy/indexed\_gzip}{corresponding
  implementation}. This is straightforward to apply for random reads
but not for random writes. Multiple reads is the only merging strategy
that works with literally no seek in the reconstructed image which
could therefore potentially use in-memory compression.

\todo{could we merge without de-compressing?}

@@

We used the blocks of the 2015 Big~Brain release with 40-micrometer
isotropic resolution available at
\url{ftp://bigbrain.loris.ca/BigBrainRelease.2015/3D\_Blocks/40um}. We
converted them to NiFTI 1.0 using Nibabel and left them uncompressed.

\todo{add slices}

@@

We used Git tag 0.1 of our \texttt{sam} library. Our experiment
scripts are available at
\url{https://github.com/big-data-lab-team/paper-sequential-split-merge/blob/master/scripts/experiment}. We
used them to split and merge using Buffered slices, Cluster reads and
Multiple reads, with 3GB, 6GB, 9GB, 12GB and 16GB of memory. We also
did 0GB of memory for Buffered slices and Cluster reads, which
triggered naive slice and block merging. We did 5 repetitions for each
memory value. Memory values were shuffled in each repetition, to
correct for potential biases coming from ordering, such as caching
effects. To improve reproducibility, we dropped the kernel page,
dentry and inode caches before each run (\texttt{echo 3 | sudo tee
  /proc/sys/vm/drop\_caches}).

@@

\subsubsection{Merging time}

% For each paragraph below, comment on overal trend and breakdown.

% Slices vs blocks
Figure~\ref{fig:slices-vs-blocks} shows the merging time for Slices vs
Blocks, on SSD and HDD. Merging with Blocks takes about twice the time
of Slices, which motivates the need for improved algorithms to merge
blocks. As expected, seek time is mainly responsible for this
difference, although differences in read times are also observed,
perhaps due to a caching effect at the disk or OS level. This behavior is
reproduced in SSD and HDD, which shows that SSDs might not be the only
solution to reduce seek time.  \todo{On figure: check that breakdowns
  add up to total time. Harmonize time formats. Harmonize colors (all
  slices in purple, all blocks in green). If time permits, try
  histogram for breakdown, on same figure than total time.}

% Buffered slices

% Cluster reads
%% ultimately fine but not stable due to discontinuities
At 3GB, Cluster reads were in case 1, which explains that more than 10
million seeks were done, almost equivalent to naive blocks. At 6, 9
and 12~GB, Cluster reads were in case 2 and at 16~GB they were in case
3. This confirms that Cluster reads are beneficial in case 2 and 3
only, consistently with the analysis in Section~\ref{sec:analysis} .

% Multiple reads
%% more stable

\begin{itemize}
\item Comment on calculation time.
\item Write time is higher than read time because use nibabel for writing but not reading.
\item It seems that the merge time is quite stable w.r.t the amount of available memory. This is good news for multiple reads. 
\end{itemize}

\begin{figure}[h]
  \centering
  \includegraphics[width=0.45\columnwidth]{figures/benchmark-buff-slices/buff-slices-breakdown-ssd.pdf}
  \hfill
    \includegraphics[width=0.45\columnwidth]{figures/benchmark--buff-slices/buff-slices-breakdown-hdd.pdf}
  \caption{Comparison between Bufered slices, naive blocks and naive slices. Left: SSD. Right: HDD.}
  % Python file: scripts/experiment/generate_avg_var_data.py
  % Gnuplot file: scripts/experiment/compare_breakdown_errbar.gnuplot
\label{fig:buffered-slices}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=0.45\columnwidth]{figures/benchmark-creads/creads-breakdown-ssd.pdf}
  \hfill
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-creads/creads-breakdown-hdd.pdf}
  \caption{Comparison between Cluster reads, naive blocks and naive slices. Left: SSD. Right: HDD.}
  % Python file: scripts/experiment/generate_avg_var_data.py
  % Gnuplot file: scripts/experiment/compare_breakdown_errbar.gnuplot
\label{fig:cluster-reads}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=0.45\columnwidth]{figures/benchmark-mreads/mreads-breakdown-ssd.pdf}
  \hfill
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-mreads/mreads-breakdown-hdd.pdf}
  \caption{Comparison between Multiple reads, naive blocks and naive slices. Left: SSD. Right: HDD.}
  % Python file: scripts/experiment/generate_avg_var_data.py
  % Gnuplot file: scripts/experiment/compare_breakdown_errbar.gnuplot
\label{fig:multiple-reads}
\end{figure}

@@

The number of seeks is reported in Figure~\ref{fig:number-of-seeks},
for all algorithms and the corresponding models
(Equations~\ref{eq:naive-blocks}
to~\ref{eq:seeks-multiple-reads}). Note the logarithmic y scale. Error
bars are not reported as numbers of seeks were constant across all
repetitions. The mean model error is \todo{x\%}, explained by the fact
that the model assumes cubic blocks while we used non-cubic ones in
the experiment. Overall, the model is validated as it correctly
explains the observations.   The measured seek times correlated with
the measured merging times ($\rho$=\todo{x}), which suggest that the
observed time reductions were indeed caused by reduced seek times.
\begin{figure}[h]
  \centering
  \includegraphics[width=\columnwidth]{figures/number-of-seeks.pdf}
  \hfill
  \caption{Number of seeks for all algorithms. y scale is logarithmic. Each algorithm is
    represented with a different color. Dark color is experimental
    value; bright color is model.}
  % Gnuplot file: scripts/number_of_seeks.gnuplot
\label{fig:number-of-seeks}
\end{figure}

% correlated to merging time

% model error

\subsection{Splitting}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.45\columnwidth]{figures/benchmark-mwrites/mwrites-breakdown-ssd.pdf}
  \hfill
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-mwrites/mwrites-breakdown-hdd.pdf}
  \caption{Comparison between Multiple writes, blocks and slices. Left: SSD. Right: HDD.}
  % Gnuplot file: scripts/blocks-slices/createboxplot.gnplt
\label{fig:multiple-writes}
\end{figure}

\newpage

@@

The complete Latex sources of this paper, including the collected data
and Gnuplot scripts used to generate the results figures are available
at \url{https://github.com/big-data-lab-team/paper-sequential-split-merge} under
GPLv3 license. Our \texttt{sam} library is at
\url{https://github.com/big-data-lab-team/sam}, under MIT license.

>>>>

% packages
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amsfonts} % for R symbol (the set of real numbers)
\usepackage{color}
\usepackage[pdftex]{graphicx}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=black,citecolor=black,filecolor=black,urlcolor=blue}
\usepackage{mathtools}
\usepackage{multirow}
\usepackage{stmaryrd} % for llbracket and rrbracket
\usepackage{subcaption}
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
\DeclarePairedDelimiter{\floor}{\lfloor}{\rfloor}

@@

This paper makes the following contributions:
\begin{itemize}
  \item We propose a set of algorithms to split and merge
    terabyte-scale 3D images from blocks or slices.
  \item We model the performance (number of seeks) of these algorithms.
  \item We evaluate our model and algorithms on a representative
    use-case involving a 1~TB image a hard drive and a solid-state
    drive.
\end{itemize}
Section~\ref{sec:algos} presents our algorithms and their complexity,
Section~\ref{sec:experiments} details the results and
Sections~\ref{sec:discussion} discusses them.\todo{Revise when outline is final.}

@@

%% \subsection{Lossless compression}

%% Lossless compression might reduce overall read and write times by
%% reducing seek time and data sizes. However, it needs to be implemented
%% on-the-fly, i.e., in chunks. See the work in~\cite{rajna2015speeding}
%% and
%% \href{https://github.com/pauldmccarthy/indexed\_gzip}{corresponding
%%   implementation}. This is straightforward to apply for random reads
%% but not for random writes. Multiple reads is the only merging strategy
%% that works with literally no seek in the reconstructed image which
%% could therefore potentially use in-memory compression.

%% \todo{could we merge without de-compressing?}

@@

We used the blocks of the 2015 Big~Brain release with 40-micrometer
isotropic resolution available at
\url{ftp://bigbrain.loris.ca/BigBrainRelease.2015/3D\_Blocks/40um}. We
converted them to NiFTI 1.0 using Nibabel and left them uncompressed.
We generated NiFTI slices of size 3850x3025x28 from the reconstructed
image, using our naive split algorithm. Slices are available at
\todo{\url{x}}.

@@

We used Git tag 0.1 of our \texttt{sam} library. Our experiment
scripts are available at
\url{https://github.com/big-data-lab-team/paper-sequential-split-merge/blob/master/scripts/experiment}. We
used them to split and merge using Buffered slices, Cluster reads and
Multiple reads, with 3~GB, 6~GB, 9~GB, 12~GB and 16~GB of
memory. Table~\ref{table:configs} shows the configurations of Cluster
reads and Multiple reads for each memory value, according to
Equations~\ref{eq:seeks-cluster-reads-1}
and~\ref{eq:seeks-multiple-reads}. For instance, Cluster reads are in
case 1 for 3~GB. We also did 0~GB of memory for Buffered slices and
Cluster reads, which triggered naive slice and block merging. We did 5
repetitions for each memory value. Memory values were shuffled in each
repetition, to correct for potential biases coming from ordering, such
as caching effects. To improve reproducibility, we dropped the kernel
page, dentry and inode caches before each run (\texttt{echo 3 | sudo
  tee /proc/sys/vm/drop\_caches}). We measured the cumulative read,
write and seek time in each run, as well as the overhead time defined
as the total time minus the sum of all other times.
\begin{table}
  \centering
\begin{footnotesize}
\begin{tabular}{|c|ccccc|}
  \hline
                 & 3~GB & 6~GB & 9~GB & 12~GB & 16~GB\\
  \hline
  Cluster  reads &  1 &  2 &  2 &  2 &  3 \\
  Multiple reads &  4 &  4 &  4 &  4 &  5\\
  \hline
\end{tabular}
\end{footnotesize}
\caption{Algorithm configurations by memory values.}
\label{table:configs}
\end{table}

@@

@@

The number of seeks is reported in Figure~\ref{fig:number-of-seeks},
for all algorithms and the corresponding models
(Equations~\ref{eq:naive-blocks}
to~\ref{eq:seeks-multiple-reads}). Note the logarithmic y scale. Error
bars are not reported as numbers of seeks were constant across all
repetitions. The mean model error is \todo{x\%}, explained by the fact
that the model assumes cubic blocks while we used non-cubic ones in
the experiment. Overall, the model is validated as it correctly
explains the observations.  
\begin{figure}[h]
  \centering
  \includegraphics[width=\columnwidth]{figures/number-of-seeks.pdf}
  \hfill
  \caption{Number of seeks for all algorithms. y scale is logarithmic. Each algorithm is
    represented with a different color. Dark color is experimental
    value; bright color is model.}
  % Gnuplot file: scripts/number_of_seeks.gnuplot
\label{fig:number-of-seeks}
\end{figure}

As expected, the difference between naive blocks and naive slices is
tremendous, in the order of 50 million seeks. For 3~GB, the number of
seeks in Cluster reads is 4 orders of magnitude higher than for
Multiple reads, and 5 orders of magnitude higher than naive
slices. This huge difference comes from the fact that at 3~GB Cluster
reads are in case 1. For 6~GB, 9~GB and 12~GB, Cluster reads are in
case 2 and the difference with Multiple reads and Buffered slices
reduces. At 16~GB, all algorithms perform the same.

\subsubsection{Merge time}

\todo{ The measured seek times correlated with
the measured merging times ($\rho$=x), which suggest that the
observed time reductions were indeed caused by reduced seek times.}

% Slices and blocks
Figure~\ref{fig:merge-time} shows the merge time for all algorithms by
memory values. Naive blocks take about 5 times longer than naive
slices to merge the data, which quantifies the effect of the problem
targeted by our algorithms. This effect appears both on HDD and SSD,
which shows that SSDs might not be the only solution to reduce seek
time. In the remainder we use naive blocks and naive slices and
references to evaluate our algorithms.

% Buffered slices
Buffered slices provide a negligible speed-up compared to naive
slices. In most cases, their memory overhead would not be worth the
time gain.

% Cluster reads
Cluster reads provide important speed ups compared to naive blocks,
both on HDD and on SSD. On SSD, they are up to 5 times faster than
naive blocks at 16~GB, and 3 on average \todo{compute the actual
  values}. \todo{On HDD, \ldots} Surprisingly, they perform
substantially faster than naive blocks even at 3~GB, while in case
1. This may be explained by the fact that the seeks required to write
incomplete block rows to the reconstructed image are shorter than the
ones for naive blocks.

% Multiple reads
Multiple reads are even faster than Cluster reads. On SSD, they are 5
times faster than naive blocks on average \todo{compute the actual
  values} ( \todo{On HDD, \ldots}). At 12~GB, the total merging time
is only 1.2 times slower than naive slices, which solves our initial
problem to a large extent. Even at 3~GB the speed-up is considerable
compared to naive blocks.

\begin{figure}[h]
  \centering
  \includegraphics[width=\columnwidth]{figures/times-hdd.pdf}\\
  \includegraphics[width=\columnwidth]{figures/times-ssd.pdf}
  \hfill
  \caption{Merge time by algorithm. Top: HDD. Bottom: SSD. Each
    algorithm is represented with a different color. Averages over 5
    repetitions. Error bars show $\pm$ 1 standard deviation. }
  % Gnuplot file: scripts/experiment/number_of_seeks.gnuplot
\label{fig:merge-time}
\end{figure}

\subsubsection{Merge time breakdown}
Figure~\ref{fig:breakdowns-ssd} 
\begin{figure}
  \begin{subfigure}[b]{\columnwidth}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-buff-slices/buff-slices-breakdown-ssd.pdf}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-buff-slices/buff-slices-breakdown-hdd.pdf}
    \caption{Buffered slices}
  \end{subfigure}
  \begin{subfigure}[b]{\columnwidth}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-creads/creads-breakdown-ssd.pdf}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-creads/creads-breakdown-hdd.pdf}
    \caption{Cluster reads}
  \end{subfigure}
  \begin{subfigure}[b]{\columnwidth}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-mreads/mreads-breakdown-ssd.pdf}
    \includegraphics[width=0.45\columnwidth]{figures/benchmark-mreads/mreads-breakdown-hdd.pdf}
    \caption{Multiple reads}
  \end{subfigure}
  \caption{Breakdown of total merge times. Left column: SSD. Right column: HDD.}
  \label{fig:breakdowns-ssd}
  % Gnuplot file: scripts/experiment/compare_breakdown_errbar.gnuplot
\end{figure}

\subsubsection{Split time}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\columnwidth]{figures/number-of-seeks.pdf}\\
  \includegraphics[width=\columnwidth]{figures/number-of-seeks.pdf}
  \hfill
  \caption{Split time by algorithm. Top: HDD. Bottom: SSD. Each
    algorithm is represented with a different color. Averages over 5
    repetitions. Error bars show $\pm$ 1 standard deviation. }
  % Gnuplot file: ???
\label{fig:split-time}
\end{figure}

@@

For more investigation and reproducibility purposes, the complete \LaTeX
sources of this paper, including the collected data and Gnuplot
scripts used to generate the results figures are available at
\url{https://github.com/big-data-lab-team/paper-sequential-split-merge}
under GPLv3 license. Our \texttt{sam} library is at
\url{https://github.com/big-data-lab-team/sam}, under MIT license.
